Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-09-20 manuscript/manuscript.tex
addressing @franknoe's change requests

Our recommended method for dimensionality reduction, TICA, is a particular implementation of the VAC.
To apply TICA, we need to compute instantaneous ($\mathbf{C}(0)$) and time-lagged ($\mathbf{C}(\tau)$) covariance matrices with elements
\begin{eqnarray}
c_{ij}(0) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t) \right\rangle_t \\
c_{ij}(\tau) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t + \tau) \right\rangle_t,
\end{eqnarray}
where $\tilde{x}_i(t)$ denotes the $i^\textrm{th}$ feature at time $t$ after the mean has been removed.
Then, we can solve the generalized eigenvalue problem
\begin{equation}
\mathbf{C}(\tau) \, \mathbf{u}_i = \mathbf{C}(0) \, \lambda_i(\tau) \, \mathbf{u}_i
\end{equation}
to obtain independent component directions $\mathbf{u}_i$ which approximate the reaction coordinates of the system,
where the pairs of eigenvalues and independent components are in descending order.

Dimensionality reduction is achieved by projecting the (mean free) features $\tilde{\mathbf{x}}(t)$
onto the leading $d$ independent components $\mathbf{U}=[\mathbf{u}_1 \dots \mathbf{u}_d]$,
\begin{equation}
\mathbf{y}(t) = \mathbf{U}_d^\top \tilde{\mathbf{x}}(t),
\end{equation}
while retaining the kinetic variance
\begin{equation}
\textrm{KV}_d = \sum\limits_{i=1}^d \lambda_i^2(\tau);
\end{equation}
the total kinetic variance is the sum of the squares of all eigenvalues.

@@

An alternative, which is much less sensitive to poor discretization,
is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint}.
HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space.
Instead, HMMs assume that there is an underlying (hidden) dynamic process which is Markovian
and gives rise to our observed data, e.g., the discretized trajectories.
This is a powerful principle as we know that there is indeed an underlying process which is Markovian:
our molecular dynamics trajectories.

To estimate an HMM, we need a spectral gap after the $m^\textrm{th}$ eigenvalue;
in practice, a timescale separation of $t_m \geq 2t_{m+1}$ is sufficient~\cite{pyemma}.
Then, we can approximate the dynamics in the observed microstates ($\mathbf{P}$) at any lag time $k\tau$ via
\begin{equation}
\mathbf{P}(k\tau) \approx \bm{\Pi}^{-1} \bm{\chi}^\top \tilde{\bm{\Pi}} \tilde{\mathbf{P}}^k(\tau) \, \bm{\chi}.
\end{equation}
Here, the $\bm{\Pi}=\left[ \pi_1,\dots,\pi_n \right]$ is a diagonal matrix of the $n$ microstates' stationary probabilities,
$\tilde{\bm{\Pi}}=\left[ \tilde{\pi}_1,\dots,\tilde{\pi}_m \right]$ is a diagonal matrix of the $m<n$ hidden states' stationary probabilities,
$\tilde{\mathbf{P}}(\tau)$ is a transition matrix between the $m<n$ hidden states at lag time $\tau$,
and $\bm{\chi}$ is an $m\times n$-dimensional row-stochastic matrix
where each row encodes the probabilities of the $n$ microstates to appear in the corresponding hidden state~\cite{noe-proj-hid-msm}.

An HMM estimation always yields a model with a small number of (hidden) states
where each state is considered to be metastable and,
thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully (see Notebook~07).
As the HMMs---like MSMs---approximate the full phase-space dynamics,
we can similarly compute physical observables from directly.

>>>>

A commonly used method for dimensionality reduction, TICA, is a particular implementation of the VAC.
To apply TICA, we need to compute instantaneous ($\mathbf{C}(0)$) and time-lagged ($\mathbf{C}(\tau)$) covariance matrices with elements
\begin{eqnarray}
c_{ij}(0) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t) \right\rangle_t \\
c_{ij}(\tau) & = & \left\langle \tilde{x}_i(t) \; \tilde{x}_j(t + \tau) \right\rangle_t,
\end{eqnarray}
where $\tilde{x}_i(t)$ denotes the $i^\textrm{th}$ feature at time $t$ after the mean has been removed.
Then, we can solve the generalized eigenvalue problem
\begin{equation}
\mathbf{C}(\tau) \, \mathbf{u}_i = \mathbf{C}(0) \, \lambda_i(\tau) \, \mathbf{u}_i \,, \quad i=1,\dots,n,
\end{equation}
to obtain independent component directions $\mathbf{u}_i$ which approximate the reaction coordinates of the system,
where the pairs of eigenvalues and independent components are sorted in descending order,
and we define a cumulative kinetic variance fraction
\begin{equation}
c_d = \frac{\sum_{i=2}^d \lambda_i^2(\tau)}{\textrm{TKV}},
\end{equation}
where
\begin{equation}
\textrm{TKV} = \sum\limits_{i=2}^n \lambda_i^2(\tau)
\end{equation}
is the total kinetic variance explained by all $n$ features.

If we further scale the independent components $\mathbf{u}_i$ by the corresponding eigenvectors $\lambda_i(\tau)$,
we obtain a \emph{kinetic map} which is the default behavior in PyEMMA.

Note, though, that TICA requires the data to be in equilibrium.
To use TICA with nonequilibrium data, we can either symmetrize the time-lagged covariance matrix $\mathbf{C}(\tau)$
or apply a Koopman reweighting~\cite{vamp-preprint}.
For short trajectories and nonequilibrium data we generally recommend to use VAMP~\cite{vamp-preprint}.

For all these approaches,
dimensionality reduction is performed by projecting the (mean free) features $\tilde{\mathbf{x}}(t)$
onto the leading $d$ independent components $\mathbf{U}_d=\left[\mathbf{u}_1 \dots \mathbf{u}_d\right]$,
\begin{equation}
\mathbf{y}(t) = \mathbf{U}_d^\top \tilde{\mathbf{x}}(t),
\end{equation}
where, in practice, $d$ is chosen such that a specific fraction of kinetic variance $c_d$ is retained (e.g., $95\%$).

@@

An alternative, which is much less sensitive to poor discretization,
is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,hmm-tutorial,jhp-spectral-rate-theory,bhmm-preprint}.
HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space.
Instead, HMMs assume that there is an underlying (hidden) dynamic process which is Markovian
and gives rise to our observed data, e.g., the discretized trajectories (see Fig.~\ref{fig:hmm-scheme}).
This is a powerful principle as we know that there is indeed an underlying process which is Markovian:
our molecular dynamics trajectories.

To estimate an HMM, we need a spectral gap after the $m^\textrm{th}$ eigenvalue;
in practice, a timescale separation of $t_m \geq 2t_{m+1}$ is sufficient~\cite{pyemma}.
Then, we can approximate the dynamics in the observed microstates ($\mathbf{P}$) at any lag time $k\tau$ via
\begin{equation}
\mathbf{P}(k\tau) \approx \bm{\Pi}^{-1} \bm{\chi}^\top \tilde{\bm{\Pi}} \tilde{\mathbf{P}}^k(\tau) \, \bm{\chi}.
\end{equation}
Here, the $\bm{\Pi}=\left[ \pi_1,\dots,\pi_n \right]$ is a diagonal matrix of the $n$ microstates' stationary probabilities,
$\tilde{\bm{\Pi}}=\left[ \tilde{\pi}_1,\dots,\tilde{\pi}_m \right]$ is a diagonal matrix of the $m<n$ hidden states' stationary probabilities,
$\tilde{\mathbf{P}}(\tau)$ is a transition matrix between the $m<n$ hidden states at lag time $\tau$,
and $\bm{\chi}$ is an $m\times n$-dimensional row-stochastic matrix
where each row encodes the emission probabilities into the $n$ microstates conditioned on being in the corresponding hidden state~\cite{noe-proj-hid-msm}.

An HMM estimation always yields a model with a small number of (hidden) states
where each state is considered to be metastable and,
thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully (see Notebook~07).
As the HMMs---like MSMs---approximate the full phase-space dynamics,
we can similarly compute the metastable kinetics, apply TPT, visualize the network, and obtain physical observables.
